The Courtroom Challenge—A Conjecture on the Unavoidable Cost of LLM Verification

### 2025-08-02 paper/main.tex
Add files via upload
Add files via upload
docs(intro, refs): cite JoT and add BibTeX entry

• Added BibTeX record for Park et al. 2024, “Judgment-of-Thought Prompting” (arXiv:2409.16635), to refs.bib.
• Inserted \cite{park2025judgment} in the Introduction.
• Recompiled main.pdf to reflect the new reference; no other content changes.

% =========================================================
\begin{document}
\maketitle
\begin{abstract}
  Large language models hallucinate; ad-hoc patches conceal the true cost of verification. 
  We therefore pose a falsifiable ultimatum: either automate validation, prove it unnecessary, or pay the explicit human-in-the-loop cost.  
  Conjecture 1 (the Courtroom Challenge). 
  For any unbounded open-world text stream, a system that lacks both (i) a persistent precedent store and (ii) a friction-priced validation loop will accumulate unbounded prequential regret.
  We formalise the state machine, cost ledger, and impossibility triangle that make this conjecture testable, and invite the community to break it.
  A 100-query toy stream illustrates how the ledger is logged\footnote{Baseline script is shown at \url{https://github.com/johnrepsys/courtroom-model}.}; no attempt is made here to optimise the validator.
  In addition, we present the Courtroom Model, a wrapper architecture that subjects every LLM claim to a perpetual cycle of precedent search, automated semantic checks, and selective escalation to an external judge—human or automated.
  Each new output is validated against a persistent vector memory, while conflicting evidence triggers the re-examination of prior “case law.”
  The challenge stands: thrive in an open world without paying the validation bill—or show why the bill is inescapable.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Large Language Models (LLMs) have demonstrated impressive capabilities in generating context-rich outputs by associating patterns from large corpora. However, this power comes with a critical limitation: LLMs often produce plausible-sounding but factually incorrect conclusions, a phenomenon known as hallucination \citep{marcus2020nextdecade,shinn2023reflexion}. Without structured verification, these errors risk being perpetuated and even amplified over time, undermining trust in LLM-generated knowledge. Addressing this issue requires a system that not only generates associations but rigorously validates and refines them.

To address these shortcomings we pose the \textit{Courtroom Challenge}: a
hybrid state machine in which every LLM claim is tested against a persistent
precedent store and—when automatic checks falter—escalated under a
tunable \textit{validation\_preference} dial (formal spec in
Section~\ref{sec:framework}).  The design rewires known parts
(generator, retrieval, human oversight) but prices each byte, cycle,
and minute of validation.  Retrieval-augmented generation (RAG) provides the raw materials
\citep{lewis2020rag,fan2024ragsurvey}.  
Recent advances—Self-RAG's self-verification \citep{asai2024selfrag}, 
Iter-RetGen's retrieval-generation synergy \citep{shao2023iterretgen}, 
MemoRAG's compressed long-term memory \citep{qian2024memorag}, and 
Chain-of-Verification RAG's answer-driven re-checks \citep{he2024covrag}—motivate 
our persistent precedent store but still validate only per answer.  
Human-alignment work such as Constitutional AI 
\citep{bai2022constitutional} motivates the escalation branch.  
Whereas MemoRAG retains past evidence without auditing it, our buffer makes
each accepted verdict both permanent and costly.
Unlike debate frameworks \citep{chen2024reconcile,madaan2023selfrefine} or formal
provers \citep{wang2025malot}, we emphasise continuous cross-session
verification: knowledge is never final, only current.

@@

Cosine and L$_2$ are special cases, but a system designer is free to
plug in any monotone metric—e.g.\ Mahalanobis distance, dot-product in a
learned embedding space, or even a hybrid score that blends semantic and
numeric keys—so long as the inequality is flipped to match the
metric's notion of proximity. Section~\ref{sec:sparse-memory-bound} analyses the cost/error trade-off
under the distance form (Lemma~\ref{lemma}), and Appendix~\ref{sec:knobs} lists this dial
among the falsifiability-critical degrees of freedom. Selecting which descriptor 
fields to embed is itself tunable; Section~\ref{sec:open-questions} (Q\ref{open-question-schema}) sketches how the Courtroom loop 
could learn or prune schema columns over time.

@@

\begin{lemma}[Interactive-Fano corollary]\label{lem:ErrLower}
Let $V$ be any validator run for $T$ rounds.
With $\kappa(\omega)$ and $\Err(\omega)$ defined in
\eqref{eq:kappaDef}–\eqref{eq:ErrDef}, write
\[
  \Err_{\mathrm{bar}}:=\E[\Err].
\]
Assume the transcript $X$ obeys the information budget
\(I(\theta;X)\le I\) under the uniform prior on an $M$-point
label set ($M\ge2$).  Then
\begin{equation}\label{eq:ErrLower-new}
  \boxed{\;
     \Err_{\mathrm{bar}}
       \;\ge\;
       \frac{\delta\,T}{4}\;
       \kappa_{\!*}\;
       \Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr),
     \quad
     \kappa_{\!*}:=\E[\kappa\mid\Err>\tfrac12\delta\kappa T]
  }
  \, .
\end{equation}
\end{lemma}

@@

\noindent
Step 2 (interactive Fano). 
Set the path-dependent loss
\[
  L^\star(m,x)
    := \mathbf{1}_{\{\Err_m(x)>\tfrac12\,\delta\,\kappa(\omega)\,T\}}
\]
and $\Delta=\tfrac12$ in Proposition 3 of
\citep{chen2024interfano}.
Inequality ($\ast$) fulfils the separation requirement
\citep[Eq.\,(5)]{chen2024interfano}.
Define
\[
  \gamma:=\frac12\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr)\in(0,\tfrac12].
\]
Then
\begin{equation}\tag{$\dagger$}
  \Pr\!\bigl[\Err(\theta,X)>\tfrac12\delta\,\kappa(\omega)T\bigr]
  \;\ge\;\gamma .
\end{equation}

\medskip\noindent
Step 3 (tail $\Rightarrow$ expectation, expanded).
With
\[
  \mathcal E:=\Bigl\{\Err>\tfrac12\delta\,\kappa T\Bigr\},
  \qquad
  \kappa_{\!*}:=\E[\kappa\mid\mathcal E],
\]
we have:
\[
\begin{aligned}
  \Err_{\mathrm{bar}}
    &=\E[\Err] \\[2pt]
    &\ge
      \E\!\Bigl[
        \tfrac12\delta\,\kappa T\;
        \mathbf{1}_{\{\Err>\tfrac12\delta\,\kappa T\}}
      \Bigr] \\[2pt]
    &=\frac{\delta T}{2}\,
      \E\!\bigl[\kappa\,\mathbf{1}_{\mathcal E}\bigr] \\[2pt]
    &=\frac{\delta T}{2}\,
      \kappa_{\!*}\;
      \Pr(\mathcal E) \\[2pt]
    &\ge \frac{\delta T}{4}\,\kappa_{\!*}\, \Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr)\;.
\end{aligned}
\]

@@

\paragraph{Discussion.}
Write the bound as  
\(\Err_{\mathrm{bar}}\!\ge(\tfrac12\delta T)\,\max\{0,\bar\kappa+\gamma-1\}\).
Because \(\gamma>0\) whenever the transcript leaks
fewer than \(\log M-\log2\) bits, any validator that
handles a non-trivial share of rounds automatically
\((\bar\kappa>1-\gamma)\)
faces a linear-in-$T$ unavoidable error floor.
Escalating a larger fraction of rounds (making $\bar\kappa$ smaller) or
relaxing the margin $\delta$ are the only ways to suppress this growth,
precisely matching the “friction buys accuracy’’ narrative.

\subsection{Sparse-memory bound}
\label{sec:sparse-memory-bound}

The following lemma gives a back-of-the-envelope bound on memory growth and justifies the choice of a retention threshold $\tau$.

\begin{lemma}
\label{lemma}
Assume (i) queries are i.i.d.\ in a bounded metric space
and (ii)  $U$ inserts a new vector only if its L$_2$ distance to every
vector in $M_t$ exceeds a threshold~$\tau$.  Then
\[
  |M_T| \;\le\;
  \bigl(\tfrac{\operatorname{diam}(Q)}{\tau}\bigr)^{d},
  \qquad
  \forall\,T\ge 0 .
\]
Hence total storage cost is $O\!\bigl(\tau^{-d}\bigr)$.
\end{lemma}

\begin{proof}[Proof sketch]
The insertion rule builds an $\tau$-packing of $Q$; a standard $\varepsilon$-net
argument bounds the packing number by $(\mathrm{diam}/\tau)^d$.
\end{proof}

@@

\item \emph{"But without numbers you cannot prove progress."} The claim is architectural, not empirical: \textsc{Claim A}: persistent precedent + adaptive friction is necessary for bounded-error knowledge bases in an open world; \textsc{Claim B}: current “bigger-is-better” pipelines violate \textsc{Claim A}. Both claims are falsifiable: show a practical system that achieves low hallucination without any external precedent store or escalation dial, and the Courtroom premise collapses. Until such a counter-example exists, producing another decimal improvement on GSM8K is beside the point.

>>>>

% =========================================================
\begin{document}
\maketitle
\begin{abstract}
Large-scale generative models hallucinate, and ad hoc patches hide the true cost of verification. We pose a falsifiable ultimatum: automate validation, prove it unnecessary, or pay the explicit human-in-the-loop bill. We conjecture that for any unbounded open-world data stream with Shannon entropy rate $h>0$, a system that lacks both a persistent precedent store and a friction-priced validation loop will accumulate unbounded prequential regret. To test this “validation trilemma” we formalise a state machine, a cost ledger, and an impossibility triangle, then invite the community to break it. A 100-query toy stream illustrates how the ledger is logged; optimising the validator is left to future work. We also introduce the Courtroom Model, a wrapper that subjects every claim—textual, visual, or otherwise—to precedent search, automated semantic checks, and selective escalation to an external judge. Each output is checked against persistent vector memory, and conflicting evidence reopens prior “case law.” The challenge stands: thrive without paying the validation bill—or show why the bill is inescapable.
\end{abstract}

\footnotetext[1]{Baseline script and toy data: \url{https://github.com/johnrepsys/courtroom-model}}

\section{Introduction}
\label{sec:intro}
Large Language Models (LLMs) have demonstrated impressive capabilities in generating context-rich outputs by associating patterns from large corpora. However, this power comes with a critical limitation: LLMs often produce plausible-sounding but factually incorrect conclusions, a phenomenon known as hallucination \citep{marcus2020nextdecade,shinn2023reflexion}. Without structured verification, these errors risk being perpetuated and even amplified over time, undermining trust in LLM-generated knowledge. Addressing this issue requires a system that not only generates associations but rigorously validates and refines them.
Throughout this paper we use the term
\emph{Courtroom} metaphorically: it denotes an epistemic due-process layer that challenges and verifies LLM outputs, not a system for practicing law or adjudicating legal disputes.

To address these shortcomings we pose the \textit{Courtroom Challenge}: a
hybrid state machine in which every LLM claim is tested against a persistent
precedent store and—when automatic checks falter—escalated under a
tunable \textit{validation\_preference} dial (formal spec in
Section~\ref{sec:framework}).  The design rewires known parts
(generator, retrieval, human oversight) but prices each byte, cycle,
and minute of validation.  Retrieval-augmented generation (RAG) provides the raw materials
\citep{lewis2020rag,fan2024ragsurvey}.  
Recent advances—Self-RAG's self-verification \citep{asai2024selfrag}, 
Iter-RetGen's retrieval-generation synergy \citep{shao2023iterretgen}, 
MemoRAG's compressed long-term memory \citep{qian2024memorag}, JoT's judgment-of-thought prompting \citep{park2025judgment}, 
and Chain-of-Verification RAG's answer-driven re-checks \citep{he2024covrag}—motivate 
our persistent precedent store but still validate only per answer.  
Human-alignment work such as Constitutional AI 
\citep{bai2022constitutional} motivates the escalation branch.  
Whereas MemoRAG retains past evidence without auditing it, our buffer makes
each accepted verdict both permanent and costly.
Unlike debate frameworks \citep{chen2024reconcile,madaan2023selfrefine} or formal
provers \citep{wang2025malot}, we emphasise continuous cross-session
verification: knowledge is never final, only current.

@@

Cosine and L$_2$ are special cases, but a system designer is free to
plug in any monotone metric—e.g.\ Mahalanobis distance, dot-product in a
learned embedding space, or even a hybrid score that blends semantic and
numeric keys—so long as the inequality is flipped to match the
metric's notion of proximity. Section~\ref{sec:sparse-memory-bound-tightness} analyses the cost/error trade-off
under the distance form, and Appendix~\ref{sec:knobs} lists this dial
among the falsifiability-critical degrees of freedom. Selecting which descriptor 
fields to embed is itself tunable; Section~\ref{sec:open-questions} (Q\ref{open-question-schema}) sketches how the Courtroom loop 
could learn or prune schema columns over time.

@@

\begin{lemma}[Interactive-Fano corollary, general prior]\label{lem:ErrLower}
Let $V$ be any validator run for $T$ rounds.
With $\kappa(\omega)$ and $\Err(\omega)$ defined in
\eqref{eq:kappaDef}-\eqref{eq:ErrDef}, write
$\Err_{\mathrm{bar}}:=\E[\Err]$.
Assume the transcript $X$ obeys the information budget
$I(\theta;X)\le I$ under an arbitrary\, prior $\mu$ on the $M$-point label set $(M\!\ge\!2)$.
Define
\[
  H_{\star}:=
  \begin{cases}
      H(\theta) & \text{(Shannon entropy)}\\[4pt]
      -\log p_{\max} & \text{(min-entropy),}
  \end{cases}
  \qquad
  p_{\max}:=\max_{m}\mu(m).
\]
Then
\begin{equation}\label{eq:ErrLower-general}
  \boxed{\,%
   \Err_{\mathrm{bar}}
   \;\ge\;
   \frac{\delta\,T}{4}\;
   \kappa_{\!*}\;
   \Bigl(1-\tfrac{I+\log 2}{H_{\star}}\Bigr),
   \quad
   \kappa_{\!*}:=
     \E\bigl[\kappa\,\bigl|\,\Err>\tfrac12\delta\kappa T\bigr]
  \,}.
\end{equation}
\end{lemma}

\begin{remark}[Entropy convention]
Throughout we write $H_{\star}$ for whichever entropy measure is appropriate
in context—Shannon entropy when a prior is specified, min-entropy when no
distributional knowledge is assumed.
Uniform priors maximise $H_{\star}$, producing the strongest (largest)
version of the bound, while skewed priors reduce $H_{\star}$ and weaken the
barrier proportionally.
\end{remark}

@@

\noindent
Step 2 (interactive Fano). 
Set the path-dependent loss
\[
  L^\star(m,x)
    := \mathbf{1}_{\{\Err_m(x)>\tfrac12\,\delta\,\kappa(\omega)\,T\}}
\]
and $\Delta=\tfrac12$ in Proposition 3 of
\citep{chen2024interfano}.
Inequality ($\ast$) fulfils the separation requirement
\citep[Eq.\,(5)]{chen2024interfano}.
Define
\[
  \gamma:=\frac12\Bigl(1-\tfrac{I+\log 2}{H_{\star}}\Bigr)\in(0,\tfrac12].
\]
Then
\begin{equation}\tag{$\dagger$}
  \Pr\!\bigl[\Err(\theta,X)>\tfrac12\delta\,\kappa(\omega)T\bigr]
  \;\ge\;\gamma .
\end{equation}

\medskip\noindent
Step 3 (tail $\Rightarrow$ expectation, expanded).
With
\[
  \mathcal E:=\Bigl\{\Err>\tfrac12\delta\,\kappa T\Bigr\},
  \qquad
  \kappa_{\!*}:=\E[\kappa\mid\mathcal E],
\]
we have:
\[
\begin{aligned}
  \Err_{\mathrm{bar}}
    &=\E[\Err] \\[2pt]
    &\ge
      \E\!\Bigl[
        \tfrac12\delta\,\kappa T\;
        \mathbf{1}_{\{\Err>\tfrac12\delta\,\kappa T\}}
      \Bigr] \\[2pt]
    &=\frac{\delta T}{2}\,
      \E\!\bigl[\kappa\,\mathbf{1}_{\mathcal E}\bigr] \\[2pt]
    &=\frac{\delta T}{2}\,
      \kappa_{\!*}\;
      \Pr(\mathcal E) \\[2pt]
    &\ge \frac{\delta T}{4}\,\kappa_{\!*}\, \Bigl(1-\tfrac{I+\log 2}{H_{\star}}\Bigr)\;.
\end{aligned}
\]

@@

\paragraph{Discussion.}
Write the bound as  
\(\Err_{\mathrm{bar}}\!\ge(\tfrac12\delta T)\,\max\{0,\bar\kappa+\gamma-1\}\).
Because \(\gamma>0\) whenever the transcript leaks
fewer than \((H_{\star}-\log 2)\) bits, any validator that
handles a non-trivial share of rounds automatically
\((\bar\kappa>1-\gamma)\)
faces a linear-in-$T$ unavoidable error floor.
Escalating a larger fraction of rounds (making $\bar\kappa$ smaller) or
relaxing the margin $\delta$ are the only ways to suppress this growth,
precisely matching the “friction buys accuracy” narrative.

\subsection{Sparse-memory bounds and tightness}
\label{sec:sparse-memory-bound-tightness}

We bound how many precedents a validator stores when it keeps only queries that
are at least a distance~$\tau$ apart.  The first lemma gives a worst-case upper
bound; the second shows that bound is tight under mild stochastic traffic.

\begin{lemma}[Sparse-memory upper bound]\label{lem:2}
Let $\mathcal Q\subset\mathbb R^{d}$ be bounded with
$\operatorname{diam}\mathcal Q=D<\infty$.
Suppose the validator maintains a set $M_t$ that starts empty and, at each
round $t$, inserts the query $q_t$ \emph{only if}
\(\min_{m\in M_{t-1}}\lVert q_t-m\rVert_2>\tau\).
Then for all $T\ge0$
\[
  |M_T|
  \;\le\;
  \Bigl(\tfrac{\operatorname{diam}\mathcal Q}{\tau}\Bigr)^{d}.
\]
Hence total storage cost is \(O\!\bigl(\tau^{-d}\bigr)\).
\end{lemma}

\begin{proof}[Proof sketch]
The rule ensures $M_T$ is a $\tau$-packing of $\mathcal Q$.
By a standard $\varepsilon$-net/volume argument, any $\tau$-packing inside a
$D$-diameter set has size at most $(D/\tau)^{d}$.
\end{proof}

\begin{lemma}[Stochastic tightness of $\tau$-packing]\label{lem:3}
Let the setting of Lemma \ref{lem:2} hold, and assume in addition that the
queries $(q_t)_{t\ge1}$ are i.i.d.\ with a density
$f\ge\lambda>0$ on $\mathcal Q$.
Define
\[
  T_0(\tau,\delta)\;=\;
  \frac{2^{d}}{\lambda\,\nu_d}\,
  \tau^{-d}\,
  \Bigl[d\log\!\bigl(D/\tau\bigr)+\log(1/\delta)\Bigr],
  \qquad
  \nu_d=\frac{\pi^{d/2}}{\Gamma(d/2+1)}.
\]
Then, for any confidence level $0<\delta<1$ and all $T\ge T_0$,
\[
  \Pr\!\Bigl[\,|M_T|\;\ge\;
        \tfrac12\,\nu_d^{-1}D^{-d}\,\tau^{-d}\Bigr]
  \;\ge\;1-\delta.
\]
In particular, under full-support i.i.d.\ traffic the memory size scales
\(\Theta(\tau^{-d})\) with high probability.
\end{lemma}

\begin{proof}[Sketch]
1.~Maximal $\tau$-packing.
Let $\{c_i\}_{i=1}^{m}\subset\mathcal Q$ be maximal with pairwise distances
$>\tau$.  By inscribing disjoint axis-aligned cubes of side $\tau/\sqrt d$ we
obtain $m\ge\alpha_d(D/\tau)^d$ with $\alpha_d=2^{-d}$.

\noindent
2.~Disjoint balls.
Set $B_i=B(c_i,\tfrac{\tau}{2})$.  The balls are disjoint and lie inside
$\mathcal Q$.

\noindent
3.~Uniform hit probability.
By the density lower bound, each ball has
$p_\tau=\lambda\nu_d2^{-d}\tau^{d}$
probability mass; the value is the same for every $i$.

\noindent
4.~Empty-ball tail.
$\Pr(B_i\text{ empty})=(1-p_\tau)^T\le e^{-p_\tau T}$
(using $\log(1-x)\le -x$).  Choosing $T\ge T_0$
forces $e^{-p_\tau T}\le\delta/m$.

\noindent
5.~Union bound and insertion rule.
With probability $\ge1-\delta$ every $B_i$ is hit at least once.
The first hit in each ball is $>\tau$ from all stored points,
hence inserted; so $|M_T|\ge m\ge\alpha_d(D/\tau)^d$.
Taking $c_d=\tfrac12\nu_d^{-1}D^{-d}\le\alpha_d$ matches the lemma's constant.
\end{proof}

\paragraph{Discussion — from tightness to the trilemma.}
Lemma~\ref{lem:2} caps memory by $O(\tau^{-d})$ for any query
sequence.  Lemma~\ref{lem:3} shows that, under full-support
i.i.d.\ traffic, the same $\Theta(\tau^{-d})$ rate is unavoidable.
Together with the Fano-style information bound in
Lemma~\ref{lem:ErrLower} we now have two orthogonal “laws of friction”:

\begin{itemize}\setlength\itemsep{2pt}
\item[\textbullet] Information ledger.
  To keep transcripts short, the validator must tolerate a
  Fano error floor.  
\item[\textbullet] Memory ledger.
  At granularity $\tau$ it must provision
  $\Theta(\tau^{-d})$ precedent cells—no compression trick beats this
  when queries explore the space fully.
\end{itemize}

\noindent
What remains open—and is precisely what the next two conjectures tackle—is a
joint lower bound that couples both ledgers plus the escalation
(human-minutes) ledger when queries arrive in an open-world, potentially
adaptive fashion.  The conjectures formulate that missing link.

@@

\item \emph{"But without numbers you cannot prove progress."} The claim is architectural, not empirical: \textsc{Claim A}: persistent precedent + adaptive friction is necessary for bounded-error knowledge bases in an open world; \textsc{Claim B}: current “bigger-is-better” pipelines violate \textsc{Claim A}. Both claims are falsifiable: show a practical system that achieves low hallucination without any external precedent store or escalation dial, and the Courtroom premise collapses. Until such a counter-example exists, producing another decimal improvement on the latest benchmark leaderboard is beside the point.
